---
title: "Diabetes Health Indicators - Modeling"
author: "Brock Akerman"
format: html
bibliography: references.bib
csl: apa.csl
---

```{r, library load, warning=FALSE, output=FALSE}
library(tidyverse)
library(dplyr)
```

```{r, read-in data, output=FALSE, warning=FALSE, error=FALSE}
#Read in the dataset
diabetes_data_raw <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

#Factorize the binary variables for better downstream analysis and remove BMI observations 50
diabetes_data_clean <- diabetes_data_raw %>%
  mutate_at(vars(Diabetes_binary, HighBP, HighChol, CholCheck, Smoker, 
                 Stroke, HeartDiseaseorAttack, PhysActivity, Fruits, 
                 Veggies, HvyAlcoholConsump, AnyHealthcare, NoDocbcCost, 
                 DiffWalk, Sex), as.factor) %>%
  filter(BMI < 50)
```



```{r, split_data}
set.seed(13579)
train <- sample(1:nrow(diabetes_data_clean), size = nrow(diabetes_data_clean)*0.7)
test <- dplyr::setdiff(1:nrow(diabetes_data_clean), train)

diabetes_train <- diabetes_data_clean[train, ]
diabetes_test <- diabetes_data_clean[test, ]
```





#### **Appendix:  Variable description**
[@cdc2015]\
**Diabetes_binary**\
_NOTE:  The options present are different from the source data._\
0 = no diabetes/prediabetes\
1 = diabetes\

**HighBP**\
0 = no high BP\
1 = high BP\

**HighChol**\
0 = no high cholesterol\
1 = high cholesterol\

**CholCheck**\
0 = no cholesterol check in 5 years\
1 = cholesterol check in 5 years\

**BMI**\
Continuous Data\

**Smoker**\
To the question:  Have you smoked at least 100 cigarettes in your entire life? [Note: 5 packs = 100 cigarettes] \
0 = no\
1 = yes\

**Stroke**\
To the question:  (Ever told) you had a stroke.\
0 = no\
1 = yes\

**HeartDiseaseorAttack**\
Coronary heart disease (CHD) or myocardial infarction (MI)\
0 = no\
1 = yes\

**PhysActivity**\
Physical activity in past 30 days (not including job)\
0 = no\
1 = yes\

**Fruits**\
Consume Fruit 1 or more times per day\
0 = no\
1 = yes\

**Veggies**\
Consume Vegetables 1 or more times per day\
0 = no\
1 = yes\

**HvyAlcoholConsump**\
(adult men >=14 drinks per week and adult women>=7 drinks per week)\
0 = no\
1 = yes\

**AnyHealthcare**\
Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc.\
0 = no\
1 = yes\

**NoDocbcCost**\
To the question:  Was there a time in the past 12 months when you needed to see a doctor but could not because of cost?\
0 = no\
1 = yes\

**GenHlth**\
To the question:  Would you say that in general your health is: scale 1-5\
1 = excellent\
2 = very good\
3 = good\
4 = fair\
5 = poor\

**MentHlth**\
Days of poor mental health scale 1-30 days\

**PhysHlth**\
physical illness or injury days in past 30 days scale 1-30\

**DiffWalk**\
Do you have serious difficulty walking or climbing stairs?\
0 = no\
1 = yes\

**Sex**\
0 = female\
1 = male\

**Age**\
_NOTE:  The options present are different from the source data._\
1 = Age 18 to 24 Respondents with reported age between 18 and 24 years (18 <= AGE <= 24)\
2 = Age 25 to 29 Respondents with reported age between 25 and 29 years (25 <= AGE <= 29)\
3 = Age 30 to 34 Respondents with reported age between 30 and 34 years (30 <= AGE <= 34)\
4 = Age 35 to 39 Respondents with reported age between 35 and 39 years (35 <= AGE <= 39)\
5 = Age 40 to 44 Respondents with reported age between 40 and 44 years (40 <= AGE <= 44)\
6 = Age 45 to 49 Respondents with reported age between 45 and 49 years (45 <= AGE <= 49)\
7 = Age 50 to 54 Respondents with reported age between 50 and 54 years (50 <= AGE <= 54)\
8 = Age 55 to 59 Respondents with reported age between 55 and 59 years (55 <= AGE <= 59)\
9 = Age 60 to 64 Respondents with reported age between 60 and 64 years (60 <= AGE <= 64)\
10 = Age 65 to 69 Respondents with reported age between 65 and 69 years (65 <= AGE <= 69)\
11 = Age 70 to 74 Respondents with reported age between 70 and 74 years (70 <= AGE <= 74)\
12 = Age 75 to 79 Respondents with reported age between 75 and 79 years (75 <= AGE <= 79)\
13 = Age 80 or older Respondents with reported age between 80 and 99 years (80 <= AGE <= 99)\

**Education**\
_NOTE:  The options present are different from the source data._\
1 = Never attended school or only kindergarten\
2 = Grades 1 through 8 (Elementary)\
3 = Grades 9 through 11 (Some high school)\
4 = Grade 12 or GED (High school graduate)\
5 = College 1 year to 3 years (Some college or technical school)\
6 = College 4 years or more (College graduate)\

**Income**\
_NOTE:  The options present are different from the source data._\
1 = Less than $10,000\
2 = Less than $15,000 ($10,000 to less than $15,000)\
3 = Less than $20,000 ($15,000 to less than $20,000)\
4 = Less than $25,000 ($20,000 to less than $25,000)\
5 = Less than $35,000 ($25,000 to less than $35,000)\
6 = Less than $50,000 ($35,000 to less than $50,000)\
7 = Less than $75,000 ($50,000 to less than $75,000)\
8 = $75,000 or more\

# REMINDER TO DROP THE LINK TO THE EDA.QMD

# References